\begin{abstract}
  A morphological tagger is a piece of computer software that provides
  complete morphological descriptions of sentences. Morphological
  taggers find applications in many NLP fields. For example, they can
  be used as a pre-processing step for syntactic parsers, in
  information retrieval and machine translation. The task of
  morphological tagging is closely related to POS tagging but
  morphological taggers provide more fine-grained morphological
  information than POS taggers. Therefore, they are often applied to
  morphologically complex languages, which extensively utilize
  inflection, derivation and compounding for encoding structural and
  semantic information. This thesis presents work on data-driven
  morphological tagging for Finnish and other morphologically complex
  languages.

  Previous work on data-driven morphological tagging for Finnish is
  scarce because of the lack of freely available manually prepared
  morphologically tagged corpora. The work presented in this thesis is
  made possible by the recently published Finnish dependency treebanks
  FinnTreeBank and Turku Dependency Treebank. Additionally, the
  Finnish open-source morphological analyzer OMorFi is extensively
  utilized in the experiments presented in the thesis.

  The thesis presents methods for improving tagging accuracy,
  estimation speed and tagging speed in presence of large structured
  morphological label sets that are typical for morphologically
  complex languages. More specifically, it presents a novel
  formulation of generative morphological taggers using weighted
  finite-state machines and applies finite-state taggers to context
  sensitive spelling correction of Finnish. The thesis also explores
  discriminative morphological tagging. It presents structured
  sub-label dependencies that can be used for improving tagging
  accuracy. Additionally, the thesis presents a cascaded variant of the
  averaged perceptron tagger. In presence of large label sets, a
  cascaded design results in substantial reduction of estimation speed
  compared to a standard perceptron tagger. Moreover, the thesis
  explores pruning strategies for perceptron taggers. Finally, the
  thesis presents the FinnPos toolkit for morphological
  tagging. FinnPos is an open-source state-of-the-art averaged
  perceptron tagger implemented by the author.
\end{abstract}
