\chapter*{List of Publications}
\addcontentsline{toc}{chapter}{List of Publications}
\markboth{List of Publications}{}
\nobibliography*
\begin{enumerate}[label=\textbf{\Roman*}]
\item\label{pub:1}\noindent\bibentry{Silfverberg2010}\\[1cm]

\item\label{pub:2}\noindent\bibentry{Silfverberg2011}\\[1cm]

\item\label{pub:3}\noindent\bibentry{Pirinen2012}\\[1cm]

\item\label{pub:4}\noindent\bibentry{Ruokolainen2014}\\[1cm]

\item\label{pub:5}\noindent\bibentry{Silfverberg2014}\\[1cm]

\item\label{pub:6}\noindent\bibentry{Silfverberg2015}
\end{enumerate}
\chapter*{Author's Contribution}
\addcontentsline{toc}{chapter}{Author's Contributions}
\markboth{Author's Contributions}{}
\subsection*{\bf Publication \ref{pub:1}: Part-of-Speech Tagging Using Parallel Weighted Finite-State Transducers}
\noindent The paper presents a weighted finite-state transducer
implementation for hidden Markov models. The present author developed
the system, performed the experiments and wrote the paper under the
supervision the second author Krister Lind\'{e}n.

\subsection*{\bf Publication \ref{pub:2}: Combining Statistical Models for {POS} Tagging using Finite-State Calculus}
\noindent The paper presents a continuation of the work in publication
\ref{pub:1}. It presents extensions to the standard hidden Markov
Model which can be conveniently implemented using weighted
finite-state transducers. The present author developed the system,
performed the experiments and wrote the paper under the supervision of
the second author Krister Lind\'{e}n.

\subsection*{\bf Publication \ref{pub:3}: Improving Finite-State Spell-Checker Suggestions with Part of Speech N-Grams}
\noindent The paper presents a spell-checker which utilizes the POS
taggers presented in publications \ref{pub:1} and \ref{pub:2} for
context modeling. The present author performed experiments together
with the first author Tommi Pirinen and participated in writing the
paper under supervision of the third author Krister Lin\'{e}n.

\subsection*{\bf Publication \ref{pub:4}: Accelerated Estimation of
  Conditional Random Fields using a Pseudo-Likelihood-inspired
  Perceptron Variant} The paper presents an alternative perceptron
inspired estimator for conditional random fields. The present author
performed the experiments, and participated in the writing of the
paper.

\subsection*{\bf Publication \ref{pub:5}: Part-of-Speech Tagging using
  Conditional Random Fields: Exploiting Sub-Label Dependencies for
  Improved Accuracy} The paper presents a system which uses
dependencies between the components of complex morphological labels
for improving tagging accuracy. The present author participated in the
design of the system and methodology of the paper, developed the
system, performed the experiments and participated in the writing of
the paper.

\subsection*{\bf Publication \ref{pub:6}: FinnPos: an open-source morphological tagging and lemmatization toolkit for Finnish} 
The paper presents a morphological tagging toolkit for Finnish and
other morphologically rich languages. The present author participated
in the design of the system and methodology of the paper, developed
the system, performed the experiments and participated in the writing
of the paper.
