\chapter{Data-Driven Lemmatization}
\label{chapter:lemmatization}

In this section, I will present the task of data driven
lemmatization. I will examine different approaches to data driven
lemmatization and present the lemmatizer used in the FinnPos toolkit
presented in Publication \ref{pub:6}.

A lemmatizer is a system that takes text as input and returns the
lemmas of words in the text. Because dictionaries and other lexical
resources often list lemmas but omit other word forms, lemmatizers are
useful for example for information extraction. They are particularly
useful for morphologically complex languages where a substantial part of
words occurring in text undergo various inflection.

Lexical resources such as dictionaries or morphological analyzers are
very helpful for the lemmatization task. In fact, lemmatization is
often seen as one of the sub tasks of morphological analysis. Another
task which is closely related to lemmatization is {\it morphological
  paradigm induction} \citep{Ahlberg2014}. Here the task is to
generate all, or a selection, of the inflectional forms of a word
form. Therefore, lemmatization can also be seen as a sub-task of
morphological paradigm induction.

\begin{table}[!htb]
\begin{center}
\begin{tabular}{l|l|l|l}
\hline
Word & Label & Translation & Lemma  \\
\hline
kissa & noun+sg+nom & a/the cat & kissa \\
sanoessa & verb+act+inf+ine & while saying (something) & sanoa \\
talossa & noun+sg+ine & in a/the house & talo
\end{tabular}
\caption{Lemmatization of three Finnish word forms ending ``-ssa''.}\label{tab:3wf}
\end{center}
\end{table}

I will treat lemmatization as a follow-up task of morphological
labeling. Therefore, the lemmatizer has access to the morphological
labels of the words in the text. The morphological label provides very
useful information for lemmatization because it can help to
disambiguate between candidate lemmas. As an example, consider the
three Finnish word form ending ``-ssa'' in Table \ref{tab:3wf}. The
different morphological analyses correspond to different ways of
forming the lemma. For example in the case of a singular inessive of a
noun (``talossa''), the lemma (``talo'') is formed by removing the
suffix ``-ssa''. If the word form is instead a nominative, the lemma
is identical to the word form. As the example shows, the morphological
label can help to rule out incorrect lemmas.

A morphological analyzer can be used for lemmatization of a
morphologically tagged text. First, analyze each word using the
morphological analyzer. This produces a set of morphological labels
and associated lemmas. Then simply pick the lemma which is associated
with the correct morphological label. As long as the morphological
label assigned to each word is also known by the morphological
analyzer, this works perfectly. Problems arise when word forms are not
recognized by the morphological analyzer or when words are assigned
morphological labels not recognized by the analyzer. There are
several approaches to solving these problems. One approach is to
utilize the morphological analyzer (for example a finite-state
analyzer) to produce a guess for a lemma even though the word form is
not recognized. The guess is based on orthographically similar words
which are recognized and can be lemmatized by the morphological
analyzer. As an example of this approach, see \cite{Linden2009} who
use finite-state algebra to transform a morphological analyzer into a
morphological guesser that can generate the lemma for words that are
not recognized by the original analyzer.

The main advantage of basing a data driven lemmatizer on an existing
morphological analyzer is that large coverage morphological analyzers
model most, if not all, morphotactic and the morphophonological
phenomena that occur in a language. Therefore, it is likely that the
analyzer recognizes a number of similar words in the inflectional paradigm
of an unknown word even though it would not recognize that specific
word form which can be utilized in lemmatization. 

Most existing work on analyzer based lemmatizers has used rather
simple statistical models. For example, \cite{Linden2009} uses plain
suffix frequencies. 

\section{Framing Lemmatization As Classification}

In contrast to lemmatizers based on morphological analyzers,
classifier based lemmatizers \cite{Chrupala2008} are learned from data
without an existing model. The general approach is based on the
observation that word forms can be transformed into lemmas using an
{\it edit script}. For example, the English noun ``dogs'' has the
lemma ``dog''. To convert ``dogs'' into ``dog'' one needs to remove
the suffix ``s''. This is a very simple example of an edit script
which I will denote $[-s \rightarrow \epsilon]$. %All edit scripts
%cannot be applied to all word forms. For example, the edit script which
%removes a final ``s'' cannot be applied on past English participle
%form ending ``ed''.
Classifier based lemmatizers frame the lemmatization task as a as
classification task. As labels, the classifier uses edit
scripts. Subsequently to labeling a word form with an edit script
class, the lemmatizer will apply the edit script thus constructing a lemma.

The main advantage of using a classifier based lemmatizer is that the
classifier can use a feature based discriminative model. In contrast
to analyzer based lemmatizers, classifier based lemmatizers can
therefore use richer information sources such as prefixes and word
shapes expressed as regular expressions \footnote{An example of a word
  shape expressions in POSIX syntax is {\tt [A-Z][a-z]+} which matches
  capitalized English words.} -- not exclusively information about
word suffixes.

Although it would be very interesting to combine these approaches, it
falls beyond the scope of this thesis. Therefore, I have used
classifier based lemmatizers. I decided upon classifier based
lemmatizers partly because the work of \cite{Linden2009} already
investigates analyzer based lemmatization for Finnish. When performing
morphological disambiguation based on the output of a morphological
analyzer, the current system does use the morphological analyzer for
lemmatization of all word forms which are recognized by the
analyzer. For all remaining words, the data driven lemmatizer is used.

In the field of morphological paradigm generation, there exists work
which in a sense combines the analyzer and classifier based
approaches, for example \citep{Hulden2014}. However, their starting
point is not a morphological analyzer. Instead a list of morphological
paradigms is used. It would be interesting to explore this but it
falls beyond the scope of the current work. Another interesting
direction for future research is joint tagging and lemmatization which
has yielded improvements both in tagging and lemmatization accuracy
\citep{Muller2015}. However, this also falls beyond the scope of the
current work.

%\begin{itemize}
%\item Classifier based: \cite{Chrupala2008}.
%\item Finite-state based: \cite{Linden2009}.
%\item Combination: \cite{Hulden2014}.
%\end{itemize}

\section{Lemmatization in FinnPos}

The classification based lemmatizer in the FinnPos toolkit reads an
input word, identifies the set of edit scripts that can be applied to
the input and scores the candidate scripts using the input form, and
its morphological label. The score is given by a feature based
classifier. Finally, the edit script that receives the highest score
is applied on the input form and the lemma is recovered.

\paragraph{Extracting Edit Scripts} Given a word form such as ``dogs''
and its lemma ``dog'', several edit scripts can be extracted. For
example, $[-s \rightarrow \epsilon]$, $[-gs \rightarrow -g]$, $[-ogs
\rightarrow -og]$. The current system extracts the shortest script
which adequately recovers the lemma. 

The FinnPos system only extracts edit scripts which delete a suffix
and appends another suffix such as the script $[-s \rightarrow
\epsilon]$. This is sufficient for Finnish where all words except
numerals exclusively exhibit inflection at the end of
words. Naturally, this would not be sufficient in general. More
general edit scripts are therefore applied with other languages, for
example by \cite{Chrupala2008}.

For morphologically complex languages, a large number of edit scripts
may be extracted from training data. For example, the Finnpos system
extracts 4835 different edit scripts for the 145953 tokens in the
training and development data of FinnTreeBank. Therefore, many of the
classes occur only a few times in the training data. This leads to data
sparsity. However, increasing the amount of the training data would
probably alleviate the problem significantly because the inventory of
inflectional paradigms, and consequently edit scripts, is finite.

\paragraph{Features for Lemmatization} For a word $w = (w_1...w_n)$ and
a morphological label $y$, the lemmatizer in the FinnPos system
currently uses the following feature templates:
\begin{itemize}
\item The word form $w$.
\item The morphological label $y$.
\item Suffixes $(w_n)$, $(w_{n-1},w_n)$, ... Up to length 10.
\item Prefixes $(w_1)$, $(w_1,w_2)$, ... Up to length 10.
\item Infixes $(w_{n-2},w_{n-1})$, $(w_{n-3},w_{n-2})$ and $(w_{n-4},w_{n-3})$.
\end{itemize}
For each feature template $f$ (except the morphological label template
$y$), FinnPos additionally uses a combination template $(f,y)$ which
captures correlations between morphological labels and the
orthographical representation of the word form.

The infix templates are useful because they model the environment
where an inflectional suffix (such as ``-ssa'') is removed and a lemma suffix
is added. They aim at preventing phonotactically impossible
combinations.

\paragraph{Estimation} The lemmatizer can be implemented using any
discriminative classifier. For example as an averaged perceptron
classifier or a logistic classifier. In the FinnPos system, the
lemmatizer is an averaged perceptron classifier.

The estimation of the lemmatizer model differs slightly from standard
averaged perceptron estimation presented in Chapter
\ref{chap:ml}. Even though the number of edit scripts can be very
large (in the order of thousands), the subset of edit scripts
applicable for any given word form is much smaller. Moreover, it is
always known in advance because it is completely determined by the
suffixes of the word form. Therefore, the classifier is only trained
to disambiguate between the possible edit scripts associated to each
word form. This speeds up estimation considerably.

\paragraph{Inference} In the FinnPos system, words which were seen
during training time, are lemmatized based on a lemma dictionary which
associates each pair of word form and morphological label with a
lemma. For words which were not seen during training or which received
a label not seen during training, are lemmatized using the data driven
lemmatizer. Additionally, a morphological analyzer can be used to
assign lemmas to those words which it recognizes.

For word forms which cannot be lemmatized using the lemma dictionary
or morphological analyzer, the data driven lemmatizer is used. For
each word form, the set of applicable edit scripts is formed and
scored. The highest scoring edit script is subsequently applied to the
word form to produce a lemma.

